Characterization and evolution of hourly extreme precipitation in France using an Explicit-Convection Regional Climate Model

### 2025-11-10 article/fig1.tex
figs independant

---
title: "Characterization and evolution of hourly extreme precipitation in France using an Explicit-Convection Regional Climate Model"

lang: fr
bibliography: bibliography.bib

format:
  pdf:
    documentclass: article
    papersize: a4
    fontsize: 11pt
    number-sections: true
    toc: false

execute:
  eval: true
  echo: false
  warning: false
  error: false

geometry: margin=1.5cm   # ajuste les marges ici
linestretch: 1.2         # espacement des lignes

header-includes:
  - \usepackage{lineno}  # package pour la numérotation des lignes
  - \linenumbers
  - \renewcommand\thelinenumber{\arabic{linenumber}} 
  - \setcounter{linenumber}{1}
  - \modulolinenumbers[5]   # affiche un numéro tous les 10 lignes

  - \usepackage{graphicx}
  - \usepackage{tabularx}
  - \usepackage{multirow}
  - \usepackage{geometry}
---

Nicolas Decoopman, Juliette Blanchet (Univ. Grenoble Alpes, CNRS, INRAE, IRD, Grenoble INP*, IGE, 38000 Grenoble, France), Antoine Blanc (RTM, ONF, 38000 Grenoble, France).

**Abstract.** Climate change is intensifying the global water cycle, with extreme precipitation events increasing in frequency and intensity. While trends in daily precipitation extremes are well-documented, sub-daily extremes—critical for flash flood risk—remain poorly characterized, especially in France, due to limited long-term, high-resolution observations. This study evaluates the ability of the AROME regional climate model (2.5 km resolution, 1959–2022), forced by ERA5 reanalysis, to reproduce observed hourly precipitation extremes and their trends, using a dense network of Météo-France stations.

We first validate AROME’s climatology against station data, showing that the model accurately captures spatial patterns of precipitation frequency, cumulative totals, and daily maxima, but systematically underestimates hourly extremes, particularly in summer. Correlation with observations is high for daily metrics (*r* > 0.9) but drops for hourly maxima (*r* = 0.6–0.7), reflecting challenges in resolving convective processes at 2.5 km resolution.

Using extreme value theory (GEV modeling), we then analyze trends in 10-year return levels for both daily and hourly extremes. At the daily scale, AROME reproduces observed positive trends in southeastern France (e.g., +20–30% in the Mercantour and Rhône Valley), consistent with previous studies and Clausius–Clapeyron scaling. At the hourly scale, trends are more heterogeneous and less robust, with significant increases in February, March, and November, but high spatial variability and low model-observation correlation in summer. Filtering for statistically significant trends improves agreement, especially in winter and late autumn, but hourly extremes remain poorly captured in spring and summer.

Our results highlight the added value of explicit-convection models for extreme precipitation studies, while underscoring their limitations for convective extremes at resolutions coarser than 1 km. The Rhône Valley emerges as a hotspot for intensifying hourly extremes, driven by increased moisture flux and orographic enhancement. This work provides the first national-scale assessment of hourly precipitation trends in France, offering new insights for flood risk management and climate adaptation strategies.

**Keywords.** Extreme precipitation, climate change, convection-permitting model, AROME, sub-daily trends, France, extreme value theory, flash floods, Clausius–Clapeyron scaling

```{python}
from __future__ import annotations

import sys
import os
from pathlib import Path
from typing import Tuple

import pandas as pd

from svgutils.transform import fromfile, SVGFigure
import cairosvg

def _to_px(value: str | None) -> float:
    """
    Convertit une longueur SVG (px, mm, cm, pt, in) en pixels (float).

    Si `value` est None ou vide, retourne 0.
    """
    if not value:
        return 0.0
    value = value.strip()
    num = ''
    unit = ''
    for ch in value:
        if ch.isdigit() or ch in '.-':
            num += ch
        else:
            unit += ch
    if not num:
        return 0.0
    numf = float(num)
    unit = unit.strip().lower()
    if unit in ('', 'px'):
        return numf
    if unit == 'mm':
        return numf * 3.779527559055  # 96 dpi
    if unit == 'cm':
        return numf * 37.79527559055
    if unit == 'in':
        return numf * 96
    if unit == 'pt':
        return numf * 1.3333333333333  # 1 pt = 1/72 in
    # Fallback: assume pixels
    return numf

def _dims(fig) -> Tuple[float, float]:
    """
    Renvoie (width, height) de `fig` en pixels.

    1) viewBox (les 2 derniers termes)
    2) attributs width/height de la racine
    3) fig.get_size()
    """
    root = fig.root  # Correction ici : on accède à la balise <svg>
    viewbox = root.get('viewBox')
    if viewbox:
        parts = [p for p in viewbox.replace(',', ' ').split() if p]
        if len(parts) == 4:
            return float(parts[2]), float(parts[3])

    # attributs width/height sur la balise <svg>
    w_attr = root.get('width')
    h_attr = root.get('height')
    if w_attr and h_attr:
        return _to_px(w_attr), _to_px(h_attr)

    # fallback
    w, h = fig.get_size()
    return _to_px(w), _to_px(h)

def assemble_vertical(arome: Path, stations: Path, legend: Path, output: Path) -> None:
    # Charger les 3 SVG
    fig_arome = fromfile(str(arome))
    fig_stations = fromfile(str(stations))
    fig_legend = fromfile(str(legend))

    # Dimensions
    w_arome, h_arome = _dims(fig_arome)
    w_stations, h_stations = _dims(fig_stations)
    w_leg, h_leg = _dims(fig_legend)

    # Canvas global
    w_maps = max(w_arome, w_stations)
    h_maps = h_arome + h_stations
    height = h_maps  # somme des deux cartes

    # Facteur d'échelle pour que la légende fasse 1.5 fois la hauteur de la carte AROME
    scale_leg = (1.5 * h_arome) / h_leg
    w_leg_scaled = w_leg * scale_leg
    h_leg_scaled = h_leg * scale_leg  # = 1.5 * h_arome

    width = w_maps + w_leg_scaled  # légende à droite

    canvas = SVGFigure(f"{width}px", f"{height}px")
    canvas.root.set('viewBox', f"0 0 {width} {height}")

    # Racines
    root_arome = fig_arome.getroot()
    root_stations = fig_stations.getroot()
    root_legend = fig_legend.getroot()

    # Redimensionnement de la légende
    root_legend.scale(scale_leg, scale_leg)

    # Positionnement
    root_stations.moveto(0, h_arome)
    # Centrage vertical de la légende sur la hauteur totale
    y_leg = (h_maps - h_leg_scaled) / 2
    root_legend.moveto(w_maps, y_leg)

    canvas.append([root_arome, root_stations, root_legend])

    os.makedirs(os.path.dirname(str(output)), exist_ok=True)
    canvas.save(str(output))
    pdf_path = str(output)[:-4] + ".pdf"
    cairosvg.svg2pdf(url=str(output), write_to=pdf_path)
    return str(output)

def assemble(arome: Path, stations: Path, legend: Path, output: Path) -> str:
    # Charger les 2 cartes
    fig_arome = fromfile(str(arome))
    fig_stations = fromfile(str(stations))

    # Dimensions via _dims existant
    w_a, h_a = _dims(fig_arome)
    w_s, h_s = _dims(fig_stations)

    # Mise à l'échelle: même hauteur que AROME
    scale_s = h_a / h_s if h_s != 0 else 1.0
    root_a = fig_arome.getroot()
    root_s = fig_stations.getroot()
    root_s.scale(scale_s)

    w_s_scaled, h_s_scaled = w_s * scale_s, h_s * scale_s

    # Canvas côte à côte
    width  = w_a + w_s_scaled
    height = max(h_a, h_s_scaled)

    canvas = SVGFigure(f"{width}px", f"{height}px")
    canvas.root.set("viewBox", f"0 0 {width} {height}")

    # Positionnement: AROME à gauche, Stations à droite
    root_a.moveto(0, 0)
    root_s.moveto(w_a, 0)

    # Pas de légende
    canvas.append([root_a, root_s])

    os.makedirs(os.path.dirname(str(output)), exist_ok=True)
    canvas.save(str(output))
    cairosvg.svg2pdf(url=str(output), write_to=str(output)[:-4] + ".pdf")
    return str(output)

import numpy as np
import matplotlib.pyplot as plt
import matplotlib.ticker as mticker

def combined_metrics_df(name_file: str):
    base_dir = Path("../outputs")         # dossier racine à explorer
    csv_paths = list(base_dir.rglob(name_file))

    frames = []
    for path in csv_paths:
        try:
            df = pd.read_csv(path)
            df["source"] = str(path.relative_to(base_dir))  # garder la provenance
            frames.append(df)
        except Exception as exc:                             # CSV illisible
            print(f"⚠️  Fichier ignoré {path}: {exc}")

    if not frames:
        raise SystemExit("Aucun metrics.csv trouvé !")

    # Concaténation puis export
    combined = pd.concat(frames, ignore_index=True)

    # --- harmonisation des échelles ---
    # cas quotidien_reduce
    mask_q = combined["source"].str.contains("quotidien_reduce", na=False)
    combined.loc[mask_q & (combined["echelle"] == "quotidien"), "echelle"] = "quotidien_reduce"

    # cas horaire_reduce
    mask_h = combined["source"].str.contains("horaire_reduce", na=False)
    combined.loc[mask_h & (combined["echelle"] == "horaire"), "echelle"] = "horaire_reduce"

    return combined

combined = combined_metrics_df("metrics.csv")
combined_signif = combined_metrics_df("metrics_signif.csv")
combined_signif["col_calculate"] = combined_signif["col_calculate"].astype(str) + "_signif"
combined_all = pd.concat([combined, combined_signif], ignore_index=True)

import math
import numpy as np
import matplotlib.pyplot as plt
import matplotlib.ticker as mticker

def plot_r_by_season(
    df,
    echelles,
    col_values,          # <-- plusieurs indicateurs (ex. ['numday', 'mean', 'mean-max'])
    nom_fichier,         # préfixe du fichier (le numéro du facet est ajouté automatiquement)
    col_plot: str="r"
):
    """
    Trace un bar‑chart groupé « r » par saison, pour plusieurs indicateurs.
    Produit un seul PNG contenant un facet‑grid (nrows × ncols).

    Parameters
    ----------
    df : pandas.DataFrame
    echelles : list[str]
        Sous-ensemble d'échelles à garder (ex. ['quotidien', 'horaire'])
    col_values : list[str]
        Liste des valeurs cibles de la colonne 'col_calculate'
    nom_fichier : str
        Nom de base du fichier PNG (sans extension)
    """
    ncols=len(col_values)

    label_map = {
        "numday": "Number of rainy days",
        "mean": "Total precipitation",
        "mean-max": "Mean of maxima",
        "z_T_p": "Significant and non-significant trend",
        "z_T_p_signif": "Significant trend"
    }

    # ---------------------------------------------
    # 0) Préparation de la grille de sous‑graphes
    # ---------------------------------------------
    n_facets = len(col_values)
    ncols = max(1, min(ncols, n_facets))
    nrows = math.ceil(n_facets / ncols)

    fig, axes = plt.subplots(
        nrows=nrows,
        ncols=ncols,
        figsize=(12, 5 * nrows),
        sharey=True
    )
    # axes → tableau 1D pour itération uniforme
    axes = np.atleast_1d(axes).flatten()

    # Pour récupérer une seule fois handles/labels de la légende
    first_handles, first_labels = None, None

    # ------------------------------------------------------------------
    # Boucle sur chaque col_value demandé → un subplot par indicateur
    # ------------------------------------------------------------------
    for ax, col_value in zip(axes, col_values):
        # --- 1) Filtrage identique ---
        sub = df.loc[
            (df["echelle"].isin(echelles)) &
            (df["col_calculate"] == col_value),
            ["season", "echelle", col_plot]
        ]
        if sub.empty:
            ax.set_visible(False)
            ax.set_title(f"Aucune donnée pour '{col_value}'")
            continue

        sub["season"] = sub["season"].str.upper()

        # --- 2) Ordre fixe ---
        ordre_total = [
            "HYDRO", 
            "SON", "DJF", "MAM", "JJA",
            "OND", "JFM", "AMJ", "JAS",
            "SEP", "OCT", "NOV", "DEC", "JAN", "FEV", 
            "MAR", "AVR", "MAI", "JUI", "JUILL", "AOU"
        ]
        saisons_pres = [s for s in ordre_total if s in sub["season"].unique()]

        # --- 3) Renommage des échelles ---
        rename_map = {
            "quotidien":        "Daily (1959-2022)",
            "quotidien_reduce": "Daily (1990-2022)",
            "horaire":          "Hourly (1990-2022)",
        }

        fr2en_month = {
            "JAN": "JAN", "FEV": "FEB", "MAR": "MAR", "AVR": "APR",
            "MAI": "MAY", "JUI": "JUN", "JUILL": "JUL",
            "AOU": "AUG", "SEP": "SEP", "OCT": "OCT",
            "NOV": "NOV", "DEC": "DEC"
        }

        # --- 4) Pivot ---
        pivot = (
            sub.pivot(index="season", columns="echelle", values=col_plot)
               .rename(columns=rename_map)
               .reindex(saisons_pres)
        )

        # --- 5) Colonnes dans l’ordre demandé ---
        ordre_cols = [
            rename_map[k] for k in ["quotidien", "quotidien_reduce", "horaire"]
            if rename_map[k] in pivot.columns
        ]
        pivot = pivot[ordre_cols]

        # --- 6) Couleurs ---
        couleurs = ["black", "dimgray", "lightgray"][:pivot.shape[1]]

        # --- 7) Bar chart groupé ---
        x = np.arange(len(pivot))
        largeur = 0.8 / pivot.shape[1]

        for i, (col, couleur) in enumerate(zip(pivot.columns, couleurs)):

            bars = ax.bar(
                x + i * largeur,
                pivot[col],
                width=largeur,
                color=couleur,
                label=col
            )

        # --- 8) Axes, grille, ticks ---

        # map index for display: prefer month-translation, else nice season or title-cased original
        display_labels = []
        for s in pivot.index:
            if s in fr2en_month:
                display_labels.append(fr2en_month[s])
            else:
                # If it's like 'JUILL' or other variants
                if s in fr2en_month:
                    display_labels.append(fr2en_month[s])
                else:
                    # fallback: capitalize first letter and lowercase the rest
                    display_labels.append(s)

        ax.set_xticks(
            x + largeur * (pivot.shape[1] - 1) / 2,
            pivot.index
        )
        ax.set_xticklabels(display_labels, rotation=45)
        ax.set_xlabel("")
        if col_plot=="me":
          ax.set_ylabel("Δ (AROME - Stations) (mm/h)")
        else:
          ax.set_ylabel(col_plot)
        ax.set_title(label_map.get(col_value, col_value))   # titre lisible pour chaque subplot)

        if col_plot=="r":
          ax.set_ylim(top=1)   # <-- Forcer ymax à 1
          ax.yaxis.set_major_locator(mticker.MultipleLocator(0.1))
          ax.yaxis.set_minor_locator(mticker.MultipleLocator(0.02))
        else:
          ax.yaxis.set_major_locator(mticker.MultipleLocator(0.5))
          ax.yaxis.set_minor_locator(mticker.MultipleLocator(0.25))

        ax.grid(axis="y", which="major", linestyle="--", alpha=0.8)
        ax.grid(axis="y", which="minor", linestyle=":",  alpha=0.5)

        # mémoriser les handles/labels la première fois
        if first_handles is None:
            first_handles, first_labels = ax.get_legend_handles_labels()

        # supprimer la légende locale seulement si elle existe
        leg = ax.get_legend()
        if leg is not None:
            leg.remove()

    # ------------------------------------------------------------------
    # 9) Légende globale (ligne en‑haut, centrée)
    # ------------------------------------------------------------------
    if first_handles:
        fig.legend(
            first_handles,
            first_labels,
            loc="lower center",
            ncol=len(first_labels),
            bbox_to_anchor=(0.5, 1.02)
        )

    fig.tight_layout()
    fig.subplots_adjust(top=0.90)  # laisser de l’espace pour la légende globale

    # ------------------------------------------------------------------
    # 10) Sauvegarde
    # ------------------------------------------------------------------
    fig.savefig(f"figures/{nom_fichier}.png", dpi=300, bbox_inches="tight")
    plt.close(fig)

def assemble_un(carte: Path, legend: Path, output: Path) -> str:
    """
    Assemble une carte SVG et sa légende à droite.
    La légende est mise à l'échelle pour avoir exactement la même hauteur que la carte.
    Exporte aussi un PDF à côté du SVG.
    """
    # Charger les SVG
    fig_map = fromfile(str(carte))
    fig_leg = fromfile(str(legend))

    # Dimensions (en px)
    w_map, h_map = _dims(fig_map)
    w_leg, h_leg = _dims(fig_leg)
    if h_map == 0 or h_leg == 0:
        raise ValueError("Hauteur nulle détectée dans la carte ou la légende.")

    # Mise à l'échelle de la légende pour égaler la hauteur de la carte
    scale_leg = h_map / h_leg
    w_leg_scaled = w_leg * scale_leg
    h_leg_scaled = h_leg * scale_leg  # = h_map

    # Canvas final
    width  = w_map + w_leg_scaled
    height = h_map
    canvas = SVGFigure(f"{width}px", f"{height}px")
    canvas.root.set("viewBox", f"0 0 {width} {height}")

    # Racines et transformations
    root_map = fig_map.getroot()
    root_leg = fig_leg.getroot()
    root_leg.scale(scale_leg, scale_leg)

    # Positionnement
    root_map.moveto(0, 0)
    # Centrage vertical robuste en cas d'arrondis
    y_leg = max(0, (h_map - h_leg_scaled) / 2.0)
    root_leg.moveto(w_map, y_leg)

    # Assemblage
    canvas.append([root_map, root_leg])

    # Sauvegardes
    os.makedirs(os.path.dirname(str(output)), exist_ok=True)
    canvas.save(str(output))
    cairosvg.svg2pdf(url=str(output), write_to=str(output)[:-4] + ".pdf")
    return str(output)

```

# Introduction and context

**Climate change** is driving a warming of the planet’s surface air, with a more pronounced increase over land than over oceans [@IPCC2021]. Global warming has reached +1.1°C worldwide, +1.7°C in metropolitan France, and +2°C in the French Alps compared to the pre-industrial era. Furthermore, the Clausius-Clapeyron relation indicates that warmer air can hold more moisture (+7% per °C) [@clapeyron1834]. Due to buoyancy (Archimedes' principle), warm air surrounded by cooler air tends to rise. As warm air ascends in the atmosphere, it undergoes adiabatic cooling, leading to the condensation of water vapor and the formation of precipitation [@meteofrance]. However, under calm conditions, the central portion of the precipitation distribution does not fully capitalize on this excess moisture. Energetic constraints (radiative balance, evaporation, ocean-air exchanges) and dynamic constraints (subsidence, synoptic winds) limit the increase in mean precipitation to only 1–3% per °C [@IPCC2021]. In contrast, during intense convective events (thunderstorms, rapid cyclogenesis), rapid ascent condenses nearly all of this surplus, causing short-duration extreme rainfall to increase by 5–8% per °C—almost matching the theoretical potential. Extreme precipitation closely follows the Clausius-Clapeyron scaling, whereas mean rainfall remains influenced by numerous other energetic and dynamic factors [@ogorman2015contrasting]. Thus, climate warming theoretically leads to an increase in extreme precipitation, though this increase varies with changes in atmospheric circulation and can be locally amplified [@blanchet2021explaining].

**Extreme precipitation events** are defined as intense rainfall over short durations (1 hour to 24 hours), corresponding to the upper tail of the precipitation intensity distribution. There is no consensus on what constitutes an "extreme" event. Some authors study precipitation intensities above the 99th percentile or seasonal/annual maxima, while others define extreme precipitation as events rarely or never encountered in a human lifetime (e.g., precipitation levels expected once every 10, 20, or 50 years). Extreme events are at the heart of climate and societal concerns, as they account for the majority of costs associated with flooding, landslides, and infrastructure failures [@IPCC_2022_WGIII]. In 2024, numerous such events made headlines, including in Nepal, Afghanistan, Central Europe, eastern Spain, and France [@wmo2024]. Notably, in June 2024, intense high-altitude rainfall contributed to major flooding in the Écrins massif [@Blanc2024]; in October 2024, over 600 mm of rain in 48 hours caused widespread flooding in the Ardèche department [@MeteoFrance2024_episodesArdeches]; and in May 2025, extremely intense but short-lived thunderstorms (locally exceeding 120 mm/h) caused extensive damage in southern Var [@MeteoFrance2025].

**Daily precipitation extremes** have increased in intensity and frequency across more than half of the world’s land regions, at a rate close to +7% per °C of warming [@IPCC2021]. Some regional studies suggest similar trends in a significant proportion of land areas [@Donat2013, @Donat2016]. In France, however, signals are far more heterogeneous and less pronounced than for temperature, with strong regional variations. In most regions, trends remain weak or non-significant, and only in certain areas—particularly the southeast—are clear signals detected. In this region, the frequency of extreme Mediterranean episodes (cumulated rainfall > 200 mm in 24 hours) has doubled between 1961 and 2020, although interannual variability remains high [@meteofrance2024_episodesMediterraneens]. The mean intensity of daily extremes has increased by +22% between 1961 and 2015 [@Ribes2019]. At the national scale, no significant trend was detectable in annual maxima of daily precipitation until the 1990s. Since then, some studies report increases of +20% to +40%, though these estimates remain sensitive to the period and analytical method [@Donat2016]. In the southeastern Alps, the increase in 20-year return level daily precipitation in autumn (1958–2017) could reach an order of magnitude comparable to its mean value (~+100%) [@blanchet2021explaining]. Projections indicate that a +4°C warming scenario could lead to an average increase of ~+15% in daily extreme rainfall across France, and up to +20% in the northern half [@soubeyroux:hal-04991790]. There remains strong spatial and interannual variability [@IPCC2021].

**Hourly extremes**, measured at 1-hour or shorter time steps, are essential for characterizing intense convective phenomena (thunderstorm downpours, stationary thunderstorms) often responsible for flash floods. Due to the lack of long, spatially dense time series, there is no systematic global analysis of sub-daily trends; available data are often sparse, short, and non-significant. Nevertheless, several regional studies detect an intensification of hourly extremes across nearly all continents, though global confidence in an overall increase remains very low [@IPCC2021]. Increases in extreme rainfall have been observed in the United States, China (summer), Australia (annually), South Africa (summer), India, Malaysia, and Italy. Depending on the method and region, studies highlight sensitivities ranging from +7% to +13% per °C—up to twice the Clausius-Clapeyron rate [@molnar2015relation]. In France, few regional studies explicitly characterize this increase in hourly extremes. Observed maximum 1-hour values now reach 40–60 mm during major Mediterranean events, compared to 30–40 mm in the 1980s–1990s [@meteofrance2024_episodesMediterraneens]. Only the study by @Berghald2025 appears to quantify trends in hourly extremes in the Alps. However, trends in hourly return levels remain weak to non-significant, with no clear spatial or seasonal coherence, contrasting with the robust signals observed at daily time scales. This suggests that hourly extremes do not yet show a clear climate signal [@Soubeyroux01022015].

**The climatological observation network** developed in the late 19th century relied on manual readings, typically once per day (around 6 UTC, which still often defines the meteorological day). A nationwide network became available from the 1950s–1960s, when coverage became sufficiently dense. Hundreds of precipitation time series exist today [@meteo-france_2020_breve_observation_classique]. At the hourly scale, the first operational automatic stations (SATIN) date back to 1967. The generalization and modernization of the real-time network occurred primarily with RADOME from 1996 onward, making hourly accumulations widely available and disseminated from the 1990s–2000s. Today, the national radar mosaic provides 5-minute data at 1-km resolution, publicly accessible in archives. ANTILOPE rainfall fields (radar + rain gauges) are operationally produced at a 15-minute resolution since 2017.

**Climate model simulations** of precipitation provide: 1) data preceding the observed record (pre-1990); and 2) spatially complete and physically consistent fields, particularly useful in poorly instrumented areas. Global Circulation Models (GCMs) were not suitable for this study, as convection is parameterized, leading to over-smoothed and poorly located hourly precipitation at resolutions of 12–200 km. The emergence of regional climate models with explicit convection (CPMs, Convection-Permitting Models) offers a unique opportunity: they realistically simulate the dynamics of intense precipitation at fine spatial and temporal scales over long periods. For this study, the model must be a Regional Climate Model (RCM, 1–4 km) with explicit convection (CPM) to explicitly represent the initiation, organization, and propagation of convective systems and their hourly extremes. Models meeting these criteria include: WRF (3 km), COSMO-CLM (2.8 km), AROME-Climate (2.5 km) [@cnrm_arome2007], ICON-LEM (2.5 km), UKCP CPM (2.2 km), CP4-Africa, and ensembles such as EURO-CP/FPS-Convection. A CP-RCM must be initialized (3D fields of temperature, humidity, pressure, wind, etc.) and laterally forced. Historically, these forcings used ERA-Interim (80 km / 6h / 1979–2019), now replaced by ERA5 (25–31 km / 1h / 1949–present), which dynamically drives a high-resolution regional model. As ERA5 is a reanalysis, the regional model it forces aims to reconstruct past hourly weather conditions, rather than producing prospective or stochastic climate simulations. This approach allows, for example, the characterization of convective environments [@ecmwf_era5_data_documentation], which can then be refined using a CPM to explicitly simulate convection. The AROME model (Application of Research to Operations at MEsoscale, 2.5 km / 1h), forced by ERA5 over the period 1959–2022, meets these criteria and was selected as the numerical model for this study. To our knowledge, it has not yet been evaluated for this purpose.

**Goal.** While trends in daily precipitation extremes are now well-documented at global and regional scales, hourly precipitation—crucial for flash floods—still suffers from limited documentation, particularly in France. Long, homogeneous observations are rare, trends are often non-significant, and targeted studies remain few. This knowledge gap stems from both the short duration of available hourly series and the physical complexity of convective processes driving these extremes. The AROME reanalysis forced by ERA5 (1959–2022) provides a unique dataset for studying hourly precipitation extremes in France. However, the validity of simulated extremes by this model has never been evaluated, either statistically or physically. Our study addresses this gap by assessing the ability of the AROME model to reproduce observed hourly extremes, comparing modeled and observed trends over several decades. It thus makes a dual contribution: 1) methodological, by evaluating new simulations against extreme observations; and 2) climatic, by documenting—for the first time—the temporal evolution of hourly extremes in France over 60 years using a validated explicit-convection model. This work sheds light on how hourly extremes evolve in a warming climate while assessing the relevance of CPMs for hydrometeorological impact studies at the local scale.

# Methodology

This study focuses on metropolitan France. The primary objective is to validate the model (AROME grid points) against observations (Météo-France stations) using simple yet representative precipitation regime indicators (correlation and bias). This step is conducted under stationary conditions, meaning long-term trends or regime shifts are not yet interpreted. Once this stationary validation is established, the analysis will be extended to non-stationary contexts (trends and variability of extremes).

## Data Used

This work utilizes precipitation data from Météo-France observations [@meteofrance2024] at daily (1959–2022) and hourly (1990–2022) time steps, as illustrated in Figure 1. Since the 2000s, homogenization procedures (PRODIGE, followed by HOMER) have been implemented in the BDClim database [@meteo-france_2020_breve_observation_classique].

<!-- FIGURE 1 -->
```{=latex}
\begin{figure}[htp]
\centering
\begin{subfigure}{0.45\textwidth}
  \centering
  \includegraphics[width=\linewidth]{../outputs/hist/dispo/quotidien/hydro/hist_len_years_cum.pdf}
  \caption{\small Stations at daily time step (n = 8198)}
\end{subfigure}%
\hfill
\begin{subfigure}{0.45\textwidth}
  \centering
  \includegraphics[width=\linewidth]{../outputs/hist/dispo/horaire/hydro/hist_len_years_cum.pdf}
  \caption{\small Stations at hourly time step (n = 2315)}
\end{subfigure}

\caption{\small Distribution of the number of hydrological years with at most 10\% missing values for Météo-France stations at the daily (1959-2022) and hourly (1990-2022) time steps.}
\end{figure}
```

The CP-RCM AROME model, recently forced by the ERA5 reanalysis, provides hourly precipitation data from 1959 to 2022 at a 2.5 km resolution over the computational domain shown in Figure 2 [@arome2014]. In metropolitan France, 87,536 grid points are generated. As part of a second internship supervised by Cécile Caillaud (Météo-France), Mathis Chevé conducted a joint study on the Clausius-Clapeyron relation, revealing that temperature trends in the model are twice as weak as observed trends. It is therefore important to note that the Clausius-Clapeyron-related component of extreme trends will theoretically be half as strong as observed trends.

<!-- FIGURE 2 -->
```{=latex}
\begin{figure}[htp]
\centering
\includegraphics[width=0.7\textwidth]{figures/espace_definition.pdf}
\caption{\small Mapping of the computational domain of the AROME numerical model.}
\end{figure}
```

At each geographical point, we calculate the proportion of missing data in the time series for each season (or month) and year. We exclude years where this proportion exceeds the 10% threshold. The number of valid remaining years is then derived. Only geographical points with at least the minimum required number of years are retained (50 years for the 1959–2022 period and 25 years for the 1990–2022 period). Subsequent analyses are restricted to this subset of stations and years.

## Indicators definitions

Seasons are defined as follows: **SON**: September (SEP), October (OCT), November (NOV), **DJF**: December (DEC), January (JAN), February (FEB), **MAM**: March (MAR), April (APR), May (MAY), **JJA**: June (JUN), July (JUL), August (AUG) and **OND**: October (OCT), November (NOV), December (DEC), **JFM**: , January (JAN), February (FEB), March (MAR), **AMJ**:  April (APR), May (MAY), June (JUN), **JAS**: July (JUL), August (AUG), September (SEP). The hydrological year (**HYDRO**) is defined as the period from 1 September of year N to 31 August of year N+1.

Using data from Météo-France observations and AROME simulations, the following metrics were computed for each AROME grid point or station, year, and month/season over the periods 1959–2022 and 1990–2022: rainy days (threshold: ≥1 mm/day), cumulative precipitation, maxima of daily and hourly precipitation.

While descriptive statistics provide useful summaries, they cannot: extrapolate beyond observed ranges, quantify uncertainty and significance or link observed trends to physical mechanisms. To address these limitations, we apply extreme value theory (EVT), specifically the Fisher–Tippett–Gnedenko theorem [@coles2001introduction]. This theorem states that, under general conditions, normalized sample maxima ($M_n$​) converge to the Generalized Extreme Value (GEV) distribution—the analogue of the Central Limit Theorem for sums. The GEV unifies three classical distributions Gumbel ($\xi \to 0$), Fréchet ($\xi > 0$) and Weibull ($\xi < 0$) where the shape parameter $\xi$ (automatically estimated) determines the tail behavior. We fit GEV distributions to maxima: first, under stationary conditions to validate model performance ; then, in a non-stationary framework, where GEV parameters vary with time. This approach enables: consistent comparison of return levels between observations and AROME, estimation of trends in extremes and linking trends to physical mechanisms.

## Modeling equations

Let $x$ denote a realization of the random variable $X$, representing the annual maximum precipitation at a given spatial point. The Generalized Extreme Value (GEV) distribution is a continuous probability distribution parameterized by the triplet $\theta = (\mu, \sigma, \xi)$ — respectively the location, scale (strictly positive), and shape—with the following cumulative distribution function:
$$
F(x;\mu ,\sigma ,\xi) = \exp\left\{-\left[1 + \xi\frac{x - \mu}{\sigma}\right]^{-1/\xi}\right\}, \quad 1 + \xi\frac{x - \mu}{\sigma} > 0
$$

### Temporal covariate

Given a time series of $n$ annual precipitation maxima for a geographical point, we assume these maxima are independent. These observations are denoted $\{x_1, x_2, \dots, x_n\}$, where each $x_i$ is an annual maximum observed at year $t_i$ and follows a GEV distribution whose parameters depend analytically on the year $i$. The year $t_i$ is transformed into a normalized covariate denoted $\tilde{t}_i$. This transformation is performed for numerical convenience but does not alter the theoretical result. We also create a temporal covariate with a breakpoint $t_+$ such that:

$$
\tilde{t}_i = \frac{t_i - t_{\min}}{t_{\max} - t_{\min}} \quad \text{with} \quad
\begin{cases}
t_{\min} = \min_i t_i \\
t_{\max} = \max_i t_i
\end{cases} \quad \text{and} \quad
\tilde{t}_{i}^\ast =
\begin{cases}
0 & \text{if } t_i < t_+ \\
\displaystyle \frac{t_i - t_+}{t_{\max} - t_+} & \text{if } t_i \ge t_+
\end{cases}
$$

This encoding allows applying a temporal slope only after the breakpoint date, with the covariate still normalized to $[0,1]$ in the post-breakpoint portion.

### Models used

Let $t \in \mathbb{N} \mid t_{\min} \leq t \leq t_{\max}$ be the temporal covariate. The stationary model is defined by $M_0(\theta_0)$ with $\theta_0 = (\mu_0, \sigma_0, \xi_0)$, where:

$$
\mu(t) = \mu_0; \quad \sigma(t) = \sigma_0; \quad \xi(t) = \xi_0.
$$

The non-stationary models are defined as follows:

$$
\begin{array}{c@{\qquad}c@{\qquad}c}
% -------- Column 1 --------
\begin{array}{c}
\begin{aligned}
M_1(\theta_1)\\[-0.2ex]
\theta_1 = (\mu_0, \mu_1, \sigma_0, \xi_0)
\end{aligned}\\[0.4ex]
\begin{cases}
\mu(t) = \mu_0 + \mu_1\, t \\
\sigma(t) = \sigma_0 \\
\xi(t) = \xi_0
\end{cases}
\end{array}
&
% -------- Column 2 --------
\begin{array}{c}
\begin{aligned}
M_2(\theta_2)\\[-0.2ex]
\theta_2 = (\mu_0, \sigma_0, \sigma_1, \xi_0)
\end{aligned}\\[0.4ex]
\begin{cases}
\mu(t) = \mu_0 \\
\sigma(t) = \sigma_0 + \sigma_1\, t \\
\xi(t) = \xi_0
\end{cases}
\end{array}
&
% -------- Column 3 --------
\begin{array}{c}
\begin{aligned}
M_3(\theta_3)\\[-0.2ex]
\theta_3 = (\mu_0, \mu_1, \sigma_0, \sigma_1, \xi_0)
\end{aligned}\\[0.4ex]
\begin{cases}
\mu(t) = \mu_0 + \mu_1\, t \\
\sigma(t) = \sigma_0 + \sigma_1\, t \\
\xi(t) = \xi_0
\end{cases}
\end{array}
\end{array}
$$

When a breakpoint $t_+$ is introduced, we denote:

$$
t^\ast = t \cdot \mathbb{1}_{t > t_+} \quad \text{with} \quad t_+ \in \mathbb{N}.
$$
The models $M_1$, $M_2$, and $M_3$ then become $M_1^\ast$, $M_2^\ast$, and $M_3^\ast$, respectively. Under this notation, $\theta_i$ becomes $\theta^\ast_i$ with $i \in \{1, 2, 3\}$.

In this study, we perform stationary and non-stationary modeling using the year as a covariate, with temporal effects on $\mu$ or $\sigma$ or both. The shape parameter $\xi$ is assumed constant. Based on the literature, we set $t_+ = 1985$. This breakpoint year is motivated by two considerations:
1. Previous studies have shown that statistically significant trends in daily precipitation extremes in southern France began in the 1980s [@Blanchet2018].
2. This year maximizes the log-likelihood [@Blanchet2018].

Using a breakpoint allows us to minimize potential biases arising from varying observation lengths while incorporating pre-1985 data to enhance the robustness of estimates.

### Return level

>>>>

\documentclass[border=4pt]{standalone}
\usepackage{graphicx}

\begin{document}

% Colonne gauche
\begin{minipage}{0.48\linewidth}
  \centering
  \includegraphics[width=\linewidth]{../outputs/hist/dispo/quotidien/hydro/hist_len_years_cum.pdf}

  {\small Stations at daily time step (n = 8198)}
\end{minipage}
\hfill
% Colonne droite
\begin{minipage}{0.48\linewidth}
  \centering
  \includegraphics[width=\linewidth]{../outputs/hist/dispo/horaire/hydro/hist_len_years_cum.pdf}

  {\small Stations at hourly time step (n = 2315)}
\end{minipage}

\end{document}
